---
title: "Data visualization assignment"
author: "Shane Grele"
date: "`r Sys.Date(2026-09-11)`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE, fig.width = 6, fig.height = 4)
```

In this exercise, you will practice basic plotting with **`ggplot2`** on
a dataset of median gene expression levels across samples from various
human tissues from the GTEx Project. The dataset (`dt`) has one row per
gene, with columns:

- `Name`: the Ensembl gene ID
- `Description`: the gene symbol
- One column per tissue containing the median expression levels (in TPM
  units)

Insert your code into the empty code blocks. Insert answers to any
comprehension questions as comments within the code blocks. You do not
need to submit the figures themselves, only the code used to produce
them. Remember to label your axes!

**Using HopGPT.** As in the last assignment: you are welcome to use it,
and wherever you do, add a couple of comment lines in that same code
block noting what you asked it and how you checked the answer.

------------------------------------------------------------------------

### Load data

Start by loading the data into R, as before.

```{r load-data}
library(tidyverse)

dt <- read_delim("~/Data/GTEx/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_median_tpm.gct.gz",
                 skip = 2, col_names = TRUE, delim = "\t")
```

Recreate the `is_mitochondrial` column from the last assignment; you
will need it in Q4 and Q5.

```{r add-mito}
dt <- dt %>%
  mutate(is_mitochondrial = str_starts(Description, "MT-"))

dt %>% count(is_mitochondrial)
```

------------------------------------------------------------------------

### Q1. Histogram

Make a histogram of gene expression in the liver. Use the argument
`breaks = seq(0, 100, 2)` to bin in units of 2 TPM, ranging from 0 to
100. Note that this cuts off the data at 100, which might be misleading
depending on what we are trying to depict! How many genes are you
leaving out?

What do you notice about the shape of the distribution? What does it
tell you about the relative expression of genes in the liver
transcriptome? Do you think this is specific to the liver? Check another
tissue, as well.

```{r q1}
#load ggplot2: 

library(ggplot2)
library(dplyr)
#plot:
p1_Liver<- dt %>% ggplot(aes(x=Liver)) + geom_histogram(breaks=seq(0, 100, 2)) + labs(x = "Liver Expression (TPM)")
p1_Liver

#genes left out: 
nrow(dt %>% filter(Liver>100)) # 594 genes have a TPM greater than 100 and are left out

#shape: skews to the right, with a much lower median than mean. 
#relative expression of liver genes: mostly low, with some very high outliers

#compare to other tissues:
p1_Testis<- dt %>% ggplot(aes(x=Testis)) + geom_histogram(breaks=seq(0,100,2)) + labs(x = "Testis Expression (TPM)")
p1_Testis

p1_Bladder<- dt %>% ggplot(aes(x=Bladder)) + geom_histogram(breaks=seq(0,100,2)) + labs(x = "Bladder Expression (TPM)")
p1_Bladder

#many other tissue types also have many genes with a very low expression score and some very highly expressed outliers, but some (such as Testis) may appear to be slightly less right-skewed

```

------------------------------------------------------------------------

### Q2. Density plot

Going back to the liver, now represent the same data with a density plot
rather than a histogram. Do not restrict the X axis. Within the same
code block, create a second plot where you remove the custom bin breaks,
but transform the expression values using `log10()`. First add a
pseudo-count of 1 to the expression values, as the log of 0 is negative
infinity.

What does the log-transformed version show you that the first plot does
not? And where on the log-transformed plot do the genes with 0 TPM end
up?

```{r q2}

# answer here

p2_Liver<- dt %>% ggplot(aes(x=Liver)) + geom_density(breaks=seq(0,100,2)) + labs(x = "Liver Expression (TPM)")
p2_Liver


p2b_Liver<- dt %>% ggplot(aes(x = log10(Liver + 1))) + geom_density() + labs(x = "Liver Expression (log(TPM))")
p2b_Liver

#The log transformed version shows the counts better, because most of the counts are close to zero. When transformed, log(0+1) is 0, so most of the zeroes show up as 0. Plotting the log10 shows that the data is right skewed, with some significantly high outliers.
```

------------------------------------------------------------------------

### Q3. Scatter plot

Make a scatter plot comparing the expression of genes in the `Liver` to
the same genes in the `Heart - Left Ventricle`. Add a pseudocount to
each and log10 transform them. Adjust the point size to 1 and the alpha
(opacity) to 0.2. Add a dashed red line along the diagonal (y = x).

What does a point far above the diagonal mean, biologically?

```{r q3}

#I asked HOPGPT how to set the y-axis since it has spaces in the name. The answer it gave was back-ticks around the column name (``).
p3_Liver_vs_Left_Ventricle <- dt %>% ggplot(aes(x=log10(Liver+1), y=log10(`Heart - Left Ventricle`+1))) +
  geom_point(alpha=0.2, size=1) + 
  geom_abline(slope=1, intercept=0, linetype= 'dashed', color="red") +
  labs(x = "Liver Expression (log(TPM))", y="Heart - Left Ventricle Expression (log(TPM))")
p3_Liver_vs_Left_Ventricle

#Biologically, any points far above the diagonal represent genes which have much greater tissue-specific expression in the Left Ventricle than in the Liver.
```

------------------------------------------------------------------------

### Q4. Colored points on scatter plot

Color the previous scatter plot points based on whether the genes are
mitochondrial vs. nuclear encoded, using the `is_mitochondrial` column
you made above.

There are only a few dozen mitochondrial genes among tens of thousands
of nuclear ones, so the colored points end up buried. Try plotting them
as a second `geom_point()` layer on top of the others, using the `data`
argument to give that layer just the mitochondrial genes:
`geom_point(data = filter(dt, is_mitochondrial), ...)`.

Where do the mitochondrial genes sit relative to the diagonal, and does
that fit what you found in the last assignment?

```{r q4}

# answer here

p4_compare_mitochondrial <- dt %>% ggplot(aes(x=log10(Liver+1), y=log10(`Heart - Left Ventricle`+1))) +
  geom_point(color= "grey", alpha=0.2, size=1) + 
  geom_point(data = filter(dt, is_mitochondrial), color= "blue", alpha=0.2, size=1) +
  geom_abline(slope=1, intercept=0, linetype= 'dashed', color="red") + labs(x = "Liver Expression (log(TPM))", y="Heart - Left Ventricle Expression (log(TPM))")
p4_compare_mitochondrial

#Many of the mitochondrial genes have a greater expression in the Left Ventricle than in the Liver, but most sit close to the dotted line, and many are highly expressed in both tissues. This concords with the results of the last assignment, which showed that both tissue types have many upregulated mitochondrial genes, as both have high energy and/or metabolic demands.
```

------------------------------------------------------------------------

### Q5. Facets

Repeat question 1 (histogram), but faceting such that you create
separate panels (arranged horizontally) for several tissues of interest:

```{r focal-tissues}
#making a new vector with focal tissues
focal_tissues <- c("Liver", "Testis", "Whole Blood", "Heart - Left Ventricle", "Brain - Cortex")

```

This requires first converting the data into "long" format with
`pivot_longer()`:

```{r pivot-long}
dt_long <- dt %>%
  select("Description", "is_mitochondrial", all_of(focal_tissues)) %>%
  pivot_longer(cols = all_of(focal_tissues), names_to = "Tissue", values_to = "TPM")
```

After creating the 5-panel plot, also split into two rows that separate
genes that are versus are not encoded by the mitochondrial genome (10
panels). Use the argument `scales = "free_y"` to deal with the fact that
there are far fewer mitochondrial-encoded than nuclear-encoded genes.

`scales = "free_y"` means the panels can no longer be compared by eye.
Write the one-sentence note you would put in a figure caption to keep a
reader from being misled.

```{r q5}

# your code for 5-panel plot

dt_long %>% filter(Tissue %in% focal_tissues) %>%
ggplot(aes(x=TPM)) + geom_histogram(breaks=seq(0,100,2)) + facet_wrap(~ Tissue) + 
  labs(x="Expression(TPM)", y= "counts")

# your code for 10-panel plot
dt_long %>% filter(Tissue %in% focal_tissues) %>%
  ggplot(aes(x=TPM)) + geom_histogram(breaks=seq(0,100,2)) + 
  facet_grid(is_mitochondrial~Tissue, scales='free_y', 
             labeller = labeller(is_mitochondrial = c(`TRUE` = "Mitochondrial", `FALSE` = "Nuclear"))) + 
               labs(x="Expression(TPM)", y= "counts")
#I asked HOPGPT how to re-label the facets in facet_grid(), given that TRUE vs. FALSE doesn't clearly indicate which genes are mitochondrial vs. nuclear


dt %>% count(is_mitochondrial) #There are 37 mitochondrially-encoded genes and 56,163 nuclear-encoded genes

# caption note
#Histograms showing the distribution of expression levels (TPM) for each gene in the GTEx dataset by tissue type for both nuclear(n=56,163), and mitochondrial(n=37) genes.
```

------------------------------------------------------------------------

Optional: Have fun and explore these data! Create your own plots and
interpret them.

```{r optional}

# answer here
```
