# Linear Regression

**Deadline:** Friday, October 9\
**Resubmission deadline:** Friday, October 30

## Assignment Overview

The goal of today's lab is to use linear regression and related statistical methods to investigate the relationship between paternal age, maternal age, and the number of de novo mutations (DNMs) in a proband (offspring). Today's assignment will build familiarity with manipulating tabular datasets containing mixed data types using the **tidyverse** in **R**. Specifically, you will import a table of de novo mutations and manipulate it to calculate the number of maternal and paternal DNMs per individual. You will then fit and interpret simple and multiple linear regression models with **stats::lm**, compare nested models with **stats::anova**, and tidy results with **broom**.

This assignment is an R Markdown notebook. Write your code in the empty code chunks and your answers where you see *Your answer:*.

## Data

Data are taken from [Halldorsson, B. V. et al. (2019). Characterizing mutagenic effects of recombination through a sequence-level genetic map. *Science*, 363(6425)](https://science.sciencemag.org/content/363/6425/eaau1043.abstract).

Read the abstract from the above paper to understand the context of the datasets you will be using. The data you need for this assignment are available from Dropbox at:

1.  Information about the number and parental origin of each de novo mutation detected in a proband (offspring): [aau1043_dnm.csv](https://www.dropbox.com/scl/fi/6e28a3dow872fi02cp537/aau1043_dnm.csv?rlkey=l3gs7fb6igff4el5ov6wg96ai&dl=0)
2.  Ages of the parents of each proband: [aau1043_parental_age.csv](https://www.dropbox.com/scl/fi/sjrq1x1g30h0j10ktxysi/aau1043_parental_age.csv?rlkey=e9g3m9iq4tfsm9ski7w4vb0bf&dl=0)

You may copy these into your submission directory (and add to your `.gitignore`).

Before beginning the assignment, take a quick look at both files (e.g., with `less -S` in Unix) to confirm their structure.

------------------------------------------------------------------------

## Getting started (R packages)

Load the tidyverse and broom packages.

```{r}
library(tidyverse)
library(broom)
```

------------------------------------------------------------------------

## Exercises

### Exercise 1: Wrangle the data

#### **Step 1.1 — Load DNMs**

Load `aau1043_dnm.csv` into a tibble.

```{r}
#Information about the number and parental origin of each de novo mutation detected in a proband (offspring):
dnms <- read_csv("~/qb26-answers/week3/aau1043_dnm.csv")
dnms
```

#### **Step 1.2 — Count DNMs by parental origin per proband**

Create a **per-proband** summary with counts of maternally and paternally inherited DNMs. Ignore DNMs without a specified parent of origin. Name the count columns `maternal_dnm` and `paternal_dnm`, as later steps use these names.

```{r}
#create a new tibble using the dnms table
per_proband_dnm_summary <- dnms %>%
  #only use DNMs with a specified parent of origin
  filter(Phase_combined %in% c("mother", "father")) %>%
  #create rows based on Proband_ID values
    group_by(Proband_id) %>%
  summarize(
  #create a maternal_dnm column that is the sum of the dnms that came from the mother in each row
    maternal_dnm = sum(Phase_combined == "mother"), 
  #create a paternal_dnm column that is the sum of the dnms that came from the father in each row
    paternal_dnm = sum(Phase_combined == "father"),
  )
per_proband_dnm_summary
```

#### **Step 1.3 — Load parental ages**

Load `aau1043_parental_age.csv`.

```{r}
#Ages of the parents of each proband:
parental_ages <- read_csv("~/qb26-answers/week3/aau1043_parental_age.csv")
parental_ages
```

#### **Step 1.4 — Merge counts with ages**

Join the two tibbles by proband ID. Name the result `merged`, as later steps use this name.

```{r}
merged <- full_join(per_proband_dnm_summary, parental_ages, by = "Proband_id")
merged
```

------------------------------------------------------------------------

### Exercise 2: Fit and interpret linear regression models with R

Use your merged data frame for the following. All plots should be clearly labeled and easily interpretable.

#### **Step 2.1 — Visualize relationships**

**2.1.1** Create a scatter plot of the count of maternal DNMs vs. maternal age → save as `ex2_a.png`

```{r}

ex2_a <- merged %>% ggplot(aes(x=Mother_age, y=maternal_dnm)) + geom_point() + labs(x="Maternal Age", y= "# of Maternal De-Novo Mutations in Proband") 
ex2_a
ggsave("ex2_a.png")
```

**2.1.2** Create a scatter plot of the count of paternal DNMs vs. paternal age → save as `ex2_b.png`

```{r}

ex2_b <- merged %>% ggplot(aes(x=Father_age, y=paternal_dnm)) + geom_point() + labs(x="Paternal Age", y= "# of Paternal De-Novo Mutations in Proband")
ex2_b
ggsave("ex2_b.png")
```

**2.1.3** Create a scatter plot of paternal age vs. maternal age → save as `ex2_c.png`

```{r}

ex2_c <- merged %>% ggplot(aes(x=Mother_age, y=Father_age)) + geom_point() + labs(x="Maternal Age", y="Paternal Age")
ex2_c
ggsave("ex2_c.png")
```

#### **Step 2.2 — OLS: maternal age vs. maternal DNMs**

Fit a simple linear regression model relating maternal age to the number of maternal de novo mutations.

```{r}

ex2_2_maternal <- lm(data=merged, maternal_dnm ~ Mother_age)
summary(ex2_2_maternal)
```

Answer the following questions:

1.  What is the "size" (i.e., slope) of this relationship? Interpret the slope in plain language. Does it match your plot?

    *Your answer: The slope is 0.3376, meaning that for every year older the mother is, the model predicts that the proband(offspring) will inherit 0.3776 additional de-novo mutations from the maternal germline.*

2.  Is the relationship significant? How do you know? Explain the p-value in plain but precise language.

    *Your answer: The relationship is significant. While the R-squared value is only \~0.23, meaning that only \~23% of the variation here is explained by the linear regression/line of best fit; the p-value is quite low, meaning that the liklihood of observing a slope this extreme or more extreme, assuming the true slope is 0, is less than 2.2E-16.*

#### **Step 2.3 — OLS: paternal age vs. paternal DNMs**

Repeat the step above but for paternal age vs. paternal DNMs.

```{r}
ex2_3_paternal <- lm(data=merged, paternal_dnm ~ Father_age)
summary(ex2_3_paternal)
```

Answer the following questions:

1.  What is the "size" (i.e., slope) of this relationship? Interpret the slope in plain language. Does it match your plot?

    *Your answer: The slope is 1.354, meaning that for every year older the father is, the model predicts that the proband(offspring) will inherit 1.354 additional de-novo mutations from the paternal germline.*

2.  Is the relationship significant? How do you know? Explain the p-value in plain but precise language.

    *Your answer: The relationship is significant. While the R-squared value is only \~0.62, meaning that only \~62% of the variation here is explained by the linear regression/line of best fit; the p-value is quite low, meaning that the liklihood of observing a slope this extreme or more extreme, assuming the true slope is 0, is less than 2.2E-16.*

#### **Step 2.4 — Predict for a 50.5-year-old father**

Use the paternal regression model to predict the expected number of paternal DNMs for a father of age 50.5. You are welcome to do this manually or using a built-in function, but show your work in the code chunk.

```{r}
ex2_4_prediction <- predict(ex2_3_paternal, newdata = data.frame(Father_age = 50.5))
ex2_4_prediction

```

*Your answer: 78.69546 de-novo paternal mutations are predicted for the proband(offspring) of a father who is 50.5 years old.*

#### **Step 2.5 — Multiple regression: correlated predictors**

Maternal DNMs arose in the mother's germline, so there is no obvious reason that the father's age should affect them.

**2.5.1** Fit a simple linear regression model relating **paternal** age to the number of **maternal** DNMs.

```{r}

ex2_5_1 <- lm(data=merged, maternal_dnm ~ Father_age)
summary(ex2_5_1)
```

1.  Is paternal age a significant predictor of maternal DNMs on its own? Using your plot from 2.1.3, explain why this might happen.

    *Your answer: While there is a 'relationship' between the two (as evidenced by the low p-value stating that the liklihood of there being a slope this extreme (0.23) or more extreme due to chance alone, assuming the true slope is 0, is very low, at 1.367E-13), this is probably because paternal age is correlated with maternal age. Maternal age is biologically relevant here, and predictive for the amount of maternal de-novo mutations.*

**2.5.2** Fit a multiple linear regression model with both maternal age and paternal age as predictors of maternal DNMs, just as we added year as a covariate in the penguin example.

```{r}

ex2_5_2 <- lm(data=merged, maternal_dnm ~ Mother_age + Father_age)
summary(ex2_5_2)
```

2.  What happened to the paternal age coefficient and its p-value?

    *Your answer: Once the mother's age is accounted for, the age of the father doesn't have a significant relationship to the amount of maternal de-novo mutations in the proband(offspring), in fact, for every year the father's age increases, the amount of maternal DNM's decreases by 0.06. However, the p-value associated with this is much higher, at 0.18.*

3.  Interpret the maternal age coefficient in plain language. How does its meaning differ from the slope you estimated in Step 2.2? (Hint: what is being held constant?)

    *Your answer: Here, paternal age is held constant for the maternal age coefficient, meaning that when paternal age is accounted for, each year the mother's age increases raises the predicted amount of de-novo maternal mutations inherited by her offspring by \~0.45, a slightly higher slope than was found in step 2.2, with a similarly low p-value of 3.41E-12.*

**2.5.3** Use `anova()` to compare your two-predictor model from 2.5.2 to your model from Step 2.2 (maternal age only).

> **Hint:** In the live coding, we used `anova()` to test whether island improved a model that already contained sex and year. The same approach works for any pair of models where the smaller model is the larger model with some terms removed (these are called "nested" models). Here, the smaller model is the one that leaves out paternal age.

```{r}

anova(ex2_5_2, ex2_2_maternal)
```

4.  Does adding paternal age improve the model? Compare the p-value from `anova()` to the p-value for paternal age in the `summary()` of the two-predictor model. Why are they the same here, when for island we needed `anova()` to get a single answer? (Hint: how many coefficients did island add to the model, and how many does paternal age add?)

    *Your answer: Adding paternal age changes the slope but does not improve the model, with a high, non-significant p-value of 0.187. The ANOVA p-value is the same (also 0.187) as the p-value for the paternal age coefficient in 2.5.2(the linear regression model with the Father_age variable added). The ANOVA is 'nested', and compares the same thing as the model in 2.5.2-- the only coefficient added is paternal age, so it asks whether paternal age improves the model. In the livecoding example with the penguins dataset, adding the 'island' variable to the linear regression model added multiple coefficients, because there were multiple islands, so making a new model accounting for 'island' made the p-value different than the ANOVA p-value, which asked whether adding 'island' improved the model.*

#### **Step 2.6 — Compare distributions of maternal vs. paternal DNMs**

Plot both distributions on the **same axes** as semi-transparent histograms; save as `ex2_d.png`.

```{r}
#put two histograms on the same ggplot, use scale_fill_manual() to create custom color fills 
ex_2_d <- ggplot() +
  geom_histogram(data = merged, aes(x = maternal_dnm, fill = "Maternal"), 
                 bins = 20, alpha = 0.5, color="grey") +
  geom_histogram(data = merged, aes(x = paternal_dnm, fill = "Paternal"), 
                 bins = 20, alpha = 0.5, color="grey") +
  scale_fill_manual(values = c("Maternal" = "pink", "Paternal" = "lightblue")) +
  labs(x = "Number of De-Novo Mutations in Proband", y = "count", fill = "Parent") 
ex_2_d
ggsave("ex2_d.png")
```

#### **Step 2.7 — Statistical test: maternal vs. paternal DNMs per proband**

We have **paired** observations per proband (maternal vs. paternal). The paired t-test assumes that the within-pair differences are approximately normally distributed.

**2.7.1** Apply a paired t-test in R using `t.test(merged$maternal_dnm, merged$paternal_dnm, paired = TRUE)`.

```{r}
ex_2_7_1 <- t.test(merged$maternal_dnm, merged$paternal_dnm, paired = TRUE)
ex_2_7_1
```

1.  What is the "size" of this relationship (i.e., the average difference in counts of maternal and paternal DNMs)? Interpret the difference in plain language. Does it match your plot?

    *Your answer: The mean difference is highly negative, meaning that per-proband(offspring), there are significantly fewer (\~39.23, p\<2.2E-16) maternal de-novo mutations on average than there are paternal de-novo mutations. It matches the histogram in step 2.6, which shows that the distribution of paternal DNMs is shifted to the right of the distribution of maternal DNMs.*

2.  Is the relationship significant? How do you know? Explain the p-value in plain but precise language.

    *Your answer: The relationship is significant. The p-value of 2.2E-16 indicates that the probablility of seeing this much of a difference (or greater) due to chance alone, assuming the true difference between maternal DNMs and paternal DNMs is 0, is less than 2.2E-16.*

**2.7.2** The paired t-test is equivalent to using the difference between the maternal and paternal DNM counts per proband as the response variable and fitting a model with only an intercept term (indicated with `1` on the right side of the model formula). Fit this model using `lm()`.

```{r}
ex_2_7_2 <- lm(data = merged, maternal_dnm-paternal_dnm ~ 1)
summary(ex_2_7_2)
```

3.  How do the results compare to the paired t-test? How would you interpret the coefficient estimate for the intercept term?

    *Your answer: The intercept, or average difference, is the same (also -39.2348, p\<2E-16), meaning that this model predicts that for a paternal DNM value of 0, the maternal DNM value is predicted to be -39.2348; in other words, the average proband is predicted to have \~39.25 fewer de-novo mutations from the maternal germline than they do DNMs from the paternal germline. The paired t-test Step 2.7.1 conveys that same information.*

------------------------------------------------------------------------

### Exercise 3: Explore a new dataset

#### **Step 3.1 — Pick a TidyTuesday dataset**

Choose a dataset from the bottom of the [TidyTuesday README](https://github.com/rfordatascience/tidytuesday) and load it.

```{r}
df_ex_3 <- read_csv("~/qb26-answers/week3/cafe.csv")
df_ex_3
```

1.  Which dataset did you choose?

    *Your answer: I chose a dataset that compared the price of a small cappuccino to the hourly wages at numerous cafes in urban, suburban, and rural environments throughout the world.*

#### **Step 3.2 — Explore and visualize**

Generate figures; save them as `ex3_<something>.png`.

```{r, fig.width=15, fig.height=8}

#wages vs. cappuccino prices colored by country
ex3_country_color<- ggplot(data=df_ex_3, aes(x = hourly_wage_gbp, y = price_gbp, color=country)) +
  geom_point() +
  labs(x = "Hourly Wage (GBP)", y = "Cappuccino Price (GBP)")
ex3_country_color
ggsave("ex3_country_color.png", width=15, height=8)

#wages vs. cappuccino prices faceted by country
ex3_country_faceted <- ggplot(df_ex_3, aes(x = hourly_wage_gbp, y = price_gbp)) +
  geom_point() +
  facet_wrap(~ country) +
  labs(x="Hourly Wage (GBP)", y="Cappuccino Price (GBP)")
ex3_country_faceted
ggsave("ex3_country_faceted.png", width=15, height=8)

#mutate the dataframe to easily look at differences between urban, suburban, and rural areas
df_ex_3_long <- df_ex_3 %>%
  pivot_longer(
    cols = c(urban, suburban, rural),
    names_to = "area",
    values_to = "present"
  ) %>%
  filter(present == TRUE)

#wages vs. cappuccino prices colored by urban vs. suburban vs. rural
ex3_area_color <- ggplot(df_ex_3_long, aes(x = hourly_wage_gbp, y = price_gbp, color=area)) +
  geom_point() + labs(x="Hourly Wage (GBP)", y="Cappuccino Price (GBP)")
ex3_area_color
ggsave("ex3_area_color.png", width=10, height=6)

#wages vs. cappuccino prices faceted by urban vs. suburban vs. rural
ex3_area_faceted <- ggplot(df_ex_3_long, aes(x = hourly_wage_gbp, y = price_gbp)) +
  geom_point() +
  facet_wrap(~ area) +
  labs(x="Hourly Wage (GBP)", y="Cappuccino Price (GBP)")
ex3_area_faceted
ggsave("ex3_area_faceted.png", width=10, height=6)

```

1.  What interesting patterns do you notice?

    *Your answer: I notice that there appears to be a correlation between wages and cappuccino prices (all reported in British Pounds), but upon first glance, there isn't any separation or clustering within country nor area type (rural vs. suburban vs. urban).*

#### **Step 3.3 — Pose and test a linear-model hypothesis**

1.  State a hypothesis that you can test with a linear model.

    *Your answer: There is a linear relationship between hourly wages and cappuccino prices, but accounting for country and area of the individual cafes will not improve the predictive power of this linear relationship.*

Fit a linear model with at least two predictors to test your hypothesis, and evaluate its fit.

```{r}
#run the linear regressin model on just one relationship, wages vs. price, to see effect of wages on price
ex_3_3_1 <- lm (data=df_ex_3_long, price_gbp ~ hourly_wage_gbp)
summary(ex_3_3_1)

#run the multiple-lm model, adding predictors, to see the effect of th
ex_3_3_2 <- lm(data = df_ex_3_long, price_gbp ~ hourly_wage_gbp + country + area)
summary(ex_3_3_2)

#run an ANOVA between the two to compare
anova(ex_3_3_1, ex_3_3_2)

```

2.  Why did you include each predictor?

    *Your answer: I included these predictors because I wanted to see the effect of both country and area on the linear model predicting the relationship between hourly wages and cappuccino prices.*

3.  Report and interpret your results. How well does the model fit?

    *Your answer: While the overall model has a lower R-squared, the p-value is very high, meaning the relationship betwen wages and prices is highly significant. Holding country and area constant improves R-squared, and still results in a low p-value for most countries and most area types, meaning that both these predictors still result in an accurate, statistically significant model The ANOVA has a very low p-value as well, meaning the predictors were shown to improve the model. However, given that there are many coefficients introduced with 'country' and three introduced with 'area', meaning that there is a risk of 'overfitting' with the coefficients when the data is clustered into this many categorical variables (countries and area types).*

------------------------------------------------------------------------

## Submission

- Your completed R Markdown notebook (`.Rmd`), with code in each chunk and your answers to all questions.
- The figures you saved (`ex2_a.png` through `ex2_d.png`, and your `ex3_` figures).

## Grading Rubric (Total = 10 points)

### **Exercise 1 — Wrangle the Data (2 points)**

- Load and inspect DNM data (**0.5 pt**)
- Create per-proband maternal and paternal DNM counts (**0.5 pt**)
- Load and inspect parental age data (**0.5 pt**)
- Join counts with ages into a merged table (**0.5 pt**)

### **Exercise 2 — Fit and Interpret Linear Models (6 points)**

- Step 2.1: Scatter plots for DNMs vs. parental age and paternal vs. maternal age (**1 pt**)
- Step 2.2: Fit and interpret maternal OLS model (**1 pt**)
- Step 2.3: Fit and interpret paternal OLS model (**0.5 pt**)
- Step 2.4: Predict paternal DNMs for age 50.5 (**0.5 pt**)
- Step 2.5: Multiple regression with both parental ages, and nested model comparison with `anova()` (**1.5 pts**)
- Step 2.6: Plot distributions of maternal vs. paternal DNMs (**0.5 pt**)
- Step 2.7: Paired t-test (`t.test` and `lm(diff ~ 1)`) and interpret results (**1 pt**)

### **Exercise 3 — Explore a New Dataset (2 points)**

- Step 3.1: Choose and document TidyTuesday dataset (**0.5 pt**)
- Step 3.2: Produce exploratory figure(s) (**0.5 pt**)
- Step 3.3: Pose and test a multiple linear regression hypothesis and interpret results (**1 pt**)

**Total Points: 10**

------------------------------------------------------------------------
